A new tidy data structure to support exploration and modeling of temporal data

### 2019-02-04 main.Rmd
small changes sect 4-5

A data pipeline describes the general flow of data through an analysis, and can generally assist in conceptualizing the process as it might be applied to a variety of problems. @unix coined the term "pipelines" in software development while developing Unix at Bell Labs. In Unix-based computer operating systems, a pipeline chains together a series of operations on the basis of their standard streams, so that the output of each program becomes the input to another. The Extract, Transform, and Load (ETL) process from recent data warehousing literature dating back to @kimball2011data outlines the workflow to prepare data for analysis, and can also be considered a data pipeline. @viewing-pipeline describes a viewing pipeline for interactive statistical graphics, that takes control of the transformation from data to plot. @xgobi, @ggobi, @orca, @plumbing and @xie2014reactive implemented data pipelines for the interactive statistical software **XGobi**, **GGobi**, **Orca**, **plumbr** and **cranvas**, respectively. The pipeline is typically described with a one way flow, from data to plot. For interactive graphics, where all plots need to be updated when a user interacts with one plot, the events typically trigger the data pipeline to be run. @xie2014reactive uses a reactive programming framework, to implement the pipeline, in which user's interactions trigger a sequence of modules to update their views, that is, practically the same as running the data pipeline producing each plot.

@@

What is notable about an effective implementation of a data pipeline is that it coordinates a user's analysis making it cleaner to follow, and permits a wider audience to focus on the data analysis without getting lost in a jungle of computational intricacies. A fluent pipeline glues tidy data and the grammar of data manipulation together. It helps (1) break up a big problem to into manageable blocks, (2) generate human readable analysis workflow, (3) avoid introducing mistakes, at least making it possible to trace them through the pipeline. New data tools developed in the R package **tsibble** [@R-tsibble] articulate the time series data pipeline, which shepherds raw temporal data through to time series analysis, and plots. More detailed explanations are given as follows.

@@

\autoref{fig:tsibble-pipeline} illustrates the distinction of a time series pipeline from a regular data pipeline. It is highly recommended to check for identical entries of key and index before constructing a tsibble. Duplicates signal the data quality issue, which would likely affect succeeding analyses and hence decision making. Analysts are encouraged to gaze at data early and reason about the process of data cleaning. When the data meets the tsibble standard, it flows neatly into the analysis stage and takes full advantage of the tsibble infrastructure.

@@

Many time operations such as lag/lead and time series models, assume an intact vector input ordered in time. Since a tsibble permits time gaps in the index, it is a good practice to check and inspect any gaps in time following the creation of a tsibble, in order to prevent inviting these avoidable errors into the analysis. The first suite of verbs (rephrasing actions performed on the object) are provided to understand and tackle implicit missing values: (1) `has_gaps()` checks if there exists time gaps; (2) `scan_gaps()` reveals all implicit missing observations; (3) `count_gaps()` summarizes the time ranges that are absent from the data; (4) `fill_gaps()` turns them into explicit ones, along with imputing by values or functions. To look into gaps over individual time periods or full-length time span, the common argument `.full` in these functions gives an option to easily switch between. The specification of `.full = TRUE` will result in fully balanced panels in other words.

@@

Besides the time gaps verbs, the **tidyverse** vocabulary is adapted and expanded to facilitate time series transformations, as listed in \autoref{tab:verb-table}. The **tidyverse** suite showcases the general-purpose verbs for effectively manipulating tabular data, for example `filter()` picks observations, `select()` picks variables, and `left_join()` joins two tables. But these verbs need handling with care when used in the time series domain. A perceivable difference is summarizing variables between data frame and tsibble using `summarize()`. The former will reduce to a single summary, whereas the latter will obtain the index and their corresponding summaries. Users who are already familiar with the **tidyverse**, will experience a gentle learning curve for mastering these verbs and glide into time series analysis with low cognitive load.

Attention has been paid to warning/error handling. The principle that underpins most verbs is a tsibble in and a tsibble out, thereby striving to maintain a valid tsibble by automatically updating index and key under the hood. If the desired temporal ordering is changed by row-wise verbs, a warning is broadcast. If a tsibble cannot be maintained in the output of a pipeline module (likely occurring to column-wise verbs), for example the index is removed by selection, an error informs users of the problem and suggests alternatives. This avoids surprising users and reminds them of the time context.

@@

Rolling window calculations are widely used techniques in time series analysis, but are not restricted to temporal applications. These operations are dependent on having an ordering, particularly time ordering for temporal data. Three common types of variations for sliding window operations are:

@@

The class of tsibble is an underlying basis of temporal data, and there is a demand for sub-classing a tsibble. For example, a fable is actually an extension of a tsibble, mentioned in \autoref{sec:ts-models}. A low-level constructor `new_tsibble()` provides a vehicle to easily create a new subclass. This new object itself is a tsibble. It perhaps needs more metadata than those of a tsibble, that gives rise to a new data extension, like prediction distributions to a fable. Tsibble verbs are also S3 generics. Developers will be able to implement these verbs for the new class if necessary.

>>>>

A data pipeline describes the flow of data through an analysis, and can generally assist in conceptualizing the process, when it is applied to a variety of problems. @unix coined the term "pipelines" in software development while developing Unix at Bell Labs. In Unix-based computer operating systems, a pipeline chains together a series of operations on the basis of their standard streams, so that the output of each program becomes the input to another. The Extract, Transform, and Load (ETL) process from recent data warehousing literature dating back to @kimball2011data outlines the workflow to prepare data for analysis, and can also be considered a data pipeline. @viewing-pipeline describes a viewing pipeline for interactive statistical graphics, that takes control of the transformation from data to plot. @xgobi, @ggobi, @orca, @plumbing and @xie2014reactive implemented data pipelines for the interactive statistical software **XGobi**, **GGobi**, **Orca**, **plumbr** and **cranvas**, respectively. The pipeline is typically described with a one way flow, from data to plot. For interactive graphics, where all plots need to be updated when a user interacts with one plot, the events typically trigger the data pipeline to be run. @xie2014reactive uses a reactive programming framework, to implement the pipeline, in which user's interactions trigger a sequence of modules to update their views, that is, practically the same as running the data pipeline producing each plot.

@@

What is notable about an effective implementation of a data pipeline is that it coordinates a user's analysis making it cleaner to follow, and permits a wider audience to focus on the data analysis without getting lost in a jungle of computational intricacies. A fluent pipeline glues tidy data and the grammar of data manipulation together. It helps (1) break up a big problem to into manageable blocks, (2) generate human readable analysis workflow, (3) avoid introducing mistakes, at least making it possible to trace them through the pipeline. New data tools developed in the R package **tsibble** [@R-tsibble] articulate the time series data pipeline, which shepherds raw temporal data through to time series analysis, and plots. More detailed explanations are given in the following sections, and the examples.

@@

\autoref{fig:tsibble-pipeline} illustrates the distinction of a time series pipeline from a regular data pipeline. It is highly recommended to check for identical entries of key and index before constructing a tsibble. Duplicates signal the data quality issue, which would likely affect subsequent analyses and hence decision making. Analysts are encouraged to gaze at data early and reason about the process of data cleaning. When the data meets the tsibble standard, it flows neatly into the analysis stage and takes full advantage of the tsibble infrastructure.

@@

Many time operations such as lag/lead and time series models, assume an intact vector input ordered in time. Since a tsibble permits time gaps in the index, it is good practice to check and inspect any gaps in time following the creation of a tsibble, in order to prevent inviting these avoidable errors into the analysis. The first suite of verbs (rephrasing actions performed on the object) are provided to understand and tackle implicit missing values: (1) `has_gaps()` checks if there exists time gaps; (2) `scan_gaps()` reveals all implicit missing observations; (3) `count_gaps()` summarizes the time ranges that are absent from the data; (4) `fill_gaps()` turns them into explicit ones, along with imputing by values or functions. To look into gaps over individual time periods or full-length time span, the common argument `.full` in these functions gives an option to easily switch between. The specification of `.full = TRUE` will result in fully balanced panels in other words.

@@

Besides the time gap verbs, the **tidyverse** vocabulary is adapted and expanded to facilitate time series transformations, as listed in \autoref{tab:verb-table}. The **tidyverse** suite showcases the general-purpose verbs for effectively manipulating tabular data, for example `filter()` picks observations, `select()` picks variables, and `left_join()` joins two tables. But these verbs need handling with care when used in the time series domain. A perceivable difference is summarizing variables between data frame and tsibble using `summarize()`. The former will reduce to a single summary, whereas the latter will obtain the index and their corresponding summaries. Users who are already familiar with the **tidyverse**, will experience a gentle learning curve for mastering these verbs and glide into time series analysis with low cognitive load.

Attention has been paid to warning and error handling. The principle that underpins most verbs is a tsibble in and a tsibble out, thereby striving to maintain a valid tsibble by automatically updating index and key under the hood. If the desired temporal ordering is changed by row-wise verbs, a warning is broadcast. If a tsibble cannot be maintained in the output of a pipeline module (likely occurring to column-wise verbs), for example the index is removed by selection, an error informs users of the problem and suggests alternatives. This avoids surprising users and reminds them of the time context.

@@

Rolling window calculations are widely used techniques in time series analysis, and often apply to other applications. These operations are dependent on having an ordering, particularly time ordering for temporal data. Three common types of variations for sliding window operations are:

@@

The class of tsibble is an underlying basis of temporal data, and there is a demand for sub-classing a tsibble. For example, a fable is actually an extension of a tsibble, mentioned in \autoref{sec:ts-models}. A low-level constructor `new_tsibble()` provides a vehicle to easily create a new subclass. This new object itself is a tsibble. It perhaps needs more metadata than those of a tsibble, that gives rise to a new data extension, like prediction distributions to a fable. Tsibble verbs are also S3 generics. Developers will be able to implement these verbs for the new class, if necessary.
